Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

The CMIP5 climate model output data is distributed across data nodes at
different climate modeling centers \citep{taylor2012overview}, but can
be accessed centrally at the World Climate Research Programme CMIP5 data
portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
register before downloading data, and some data are restricted to
non-commercial use. There is a separate file for each combination of
climate model, experiment, modeling realm (e.g., atmosphere, ocean),
variable, time step, and ensemble member
\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the
output is further split across multiple files for specific year ranges
(e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each
file's name includes the output variable, climate model, experiment, and
ensemble member for the simulation \citep{taylor2010cmip5}.

@@

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
\label{fig:timeseries}
\end{figure}

CMIP5 data are available as netCDF files, for which base R import
functions do not exist. There are, however, a few R packages that can be
used to work with the netCDF file format used for CMIP5 files. Older
packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
the newer netCDF version 4 released in 2008 and are no longer available
through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
version 4 and netCDF's older version 3. While climate model output data
for CMIP5 is required to conform with the earlier version (version 3)
\citep{taylor2010cmip5}, it is safer to write code using functions that
can be used with either version, in case future phases of CMIP do not
require files to conform with netCDF version 3.

@@

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special \code{cmip5data}
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data \citep{RCMIP5}. While the \code{loadCMIP5} function does
successfully load daily data as a \code{cmip5data} object, most of the
methods for this object type do not do anything meaningful for daily
data. The package's \code{getFileInfo} function, however, will work with
CMIP5 files of any time step; this function identifies all CMIP5 files
in a directory and creates a dataframe with information parsed from the
file name. As a note, the \code{get.split.filename.cmip5} function in
the \pkg{ncdf4.helper} package similarly can be used to parse
information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. The package then uses the
\code{models2wux} function to read in climate model output netCDF files
and convert them to ``WUX'' dataframes, which can be used by other
functions in the package. While this function can input climate model
output with daily time steps (the ``what.timesteps'' element of the
\code{modelinput} list input must be set to ``daily''), the function
aggregates this data to a monthly or less frequent (e.g., seasonal)
aggregation when creating the WUX dataframe. Therefore, while this
package provides very useful functionality for working with averaged
output of daily climate model output data, it cannot easily be used to
identify and characterize multi-day extreme events like heat waves.

@@

When using this package, CMIP5 climate model output data requires some
pre-processing. Data will need to be saved in a specific format, with
files stored in a specific directory structure. Full details of the
required file and directory structure are provided in the package's
``futureheatwaves'' vignette, while tips and an R script for conducting
this processing starting from CMIP5 netCDF files are given in the
``starting\_from\_netcdf'' vignette.

@@

However, this is not the only accepted heat wave definition in the
scientific literature. A variety of different heat wave definitions have
been used to identify heat waves in a time series of temperature data
\citep{smith2013heat}, and the choice of heat wave definitions can
influence both projected heat wave trends \citep{smith2013heat} and
estimates of health risks during events
\citep{chen2015influence, kent2014heat}. Further, other types of extreme
events will be defined differently than heat waves (for example, frost
day spells may be defined as one or more days with temperature at or
below 32\textsuperscript{o}F / 0\textsuperscript{o}C).

Therefore, this package allows the user to extensively customize the
definition used to identify extreme events. Users can write a custom R
function with either a different heat wave definition (see
\citet{smith2013heat} and \citet{kent2014heat} for listings of
definitions used in scientific studies) or with a definition appropriate
for a different type of extreme event (e.g., one or more days at or
below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day
spells). For heat wave identification, researchers might want to use a
specific definition, for example, because it matches the definition used
by local health officials to declare heat wave warnings or, in the case
of health impact assessments, to match with a definition used in an
epidemiological study. For other extreme events, a heat wave definition
likely will not be applicable and so a customized definition is
necessary.

Three components of the extreme event definition can be easily
customized in the \code{gen\_hw\_set} function call, without creating a
new R function to use to identify heat waves. First, often users wish to
define an extreme event based on conditions that are rare in the study
location \citep{IPCCch1}. In this case, the percentile of the variable
of interest required for an extreme event can be changed using the
\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
option could be used to identify heat waves based on the
99\textsuperscript{th} percentile, rather than the
98\textsuperscript{th} percentile, of a city's year-round temperature.
Second, the user can change the number of days used in the heat wave
definition using the \code{numDays} argument in the \code{gen\_hw\_set}
function (the default is \samp{numDays = 2}, corresponding to two or
more days). Combined, these two customization choices allow the user to
identify heat waves using many of the heat wave definitions used in
previous climate and health research-- for example, 9 of the 16 heat
wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
the 9 that could be fit{]}) could be fit using different combinations of
these two options for specifying threshold percentile and number of
days. Third, some extreme events like cold waves and frost day spells
are defined as a certain number of days below, rather than above, a
threshold. While the default is to identify events by searching for days
above a threshold, this behavior can be changed with the
\code{above\_threshold = FALSE} argument in the
\code{gen\_hw\_set function}.

Beyond these simpler options, the customization of the event definition
is even more extensive as one has the option of writing a custom R
function. This functionality allows the user to use heat wave
definitions that either require a number of days above an absolute
threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
that require a combination of thresholds to be met (e.g., maximum daily
temperature one threshold every day of the heat wave and above a higher
threshold for a certain number of days;
\citet{kent2014heat, peng2011toward}). To use a customized event
definition, the user can write and load an R function that implements
the definition. The custom function is passed to the \code{gen\_hw\_set}
function using the \code{IDheatwavesFunction} argument. To work
correctly, this custom function must allow only specific inputs and
generate only specific outputs. Details about this required structure
are provided in the \pkg{futureheatwaves} package vignette. To increase
processing speed when identifying extreme events, we coded parts of the
default event definition function in C++ and synced it with R using the
\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
strategy for custom heat wave definitions, especially if processing a
large number of climate projection files.

Finally, extreme events are often based on conditions that are rare for
a specific location using location-specific relative thresholds. These
thresholds are often defined for climate impact studies based on a
variable's distribution at that location in present-day or historical
data. However, it can be interesting to explore trends in extreme events
under climate change if extreme events are identified based on variable
distributions during the projection period, or another future period. It
is possible to specify the time period to use when determining a
relative threshold for an event definition using the
\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.

\subsection{Customizing dates for projections and reference
temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}

Depending on the climate model output data used and the research
question, different projection periods may be of interest. For example,
one researcher may wish to explore extreme events in the 2050--2075
period, while another may be interested in 2090--2100. The projection
date range can be changed by specifying starting and ending years in the
\code{projectionBoundaries} argument of the \code{gen\_hw\_set}
function.

Further, some of the event characteristics (e.g.,
\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are
calculated based on relative temperature, which is a measure of how the
value of the variable of interest during an extreme event compares to
the typical distribution of that variable at that location. These
characteristics are measured for each of the extreme events identified
by the \code{gen\_hw\_set} function by taking the absolute value of the
variable during the event (e.g., average temperature during the heat
wave is 90\textsuperscript{o}F ~32.2\textsuperscript{o}C) and comparing
it to the location's typical variable distribution. This process
generates relative measures of how intense the event is compared to what
is normal in that location (e.g., 90\textsuperscript{o}F
~32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of
year-round temperatures in the location).

>>>>

The CMIP5 climate model output data are distributed across data nodes at
different climate modeling centers \citep{taylor2012overview}, but can
be accessed centrally at the World Climate Research Programme CMIP5 data
portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
register before downloading data, and some data are restricted to
non-commercial use. There is a separate file for each combination of
climate model, experiment, modeling realm (e.g., atmosphere, ocean),
variable, time step, and ensemble member
\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the
output is further split across multiple files for specific year ranges
(e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each
file's name includes the output variable, climate model, experiment, and
ensemble member for the simulation \citep{taylor2010cmip5}.

@@

CMIP5 data are available as netCDF files, for which base R import
functions do not exist. There are, however, a few R packages that can be
used to work with the netCDF file format used for CMIP5 files. Older
packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
the newer netCDF version 4 released in 2008 and are no longer available
through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
version 4 and netCDF's older version 3. While climate model output data
for CMIP5 are required to conform with the earlier version (version 3)
\citep{taylor2010cmip5}, it is safer to write code using functions that
can be used with either version, in case future phases of CMIP do not
require files to conform with netCDF version 3.

@@

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075. Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member. Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
\label{fig:timeseries}
\end{figure}

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special \code{cmip5data}
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data \citep{RCMIP5}. While the \code{loadCMIP5} function does
successfully load daily data as a \code{cmip5data} object, most of the
methods for this object type do not do anything meaningful for daily
data. The package's \code{getFileInfo} function, however, will work with
CMIP5 files of any time step; this function identifies all CMIP5 files
in a directory and creates a data frame with information parsed from the
file name. As a note, the \code{get.split.filename.cmip5} function in
the \pkg{ncdf4.helper} package similarly can be used to parse
information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. The package then uses the
\code{models2wux} function to read in climate model output netCDF files
and convert them to ``WUX'' data frames, which can be used by other
functions in the package. While this function can input climate model
output with daily time steps (the ``what.timesteps'' element of the
\code{modelinput} list input must be set to ``daily''), the function
aggregates this data to a monthly or less frequent (e.g., seasonal)
aggregation when creating the WUX data frame. Therefore, while this
package provides very useful functionality for working with averaged
output of daily climate model output data, it cannot easily be used to
identify and characterize multi-day extreme events like heat waves.

@@

When using this package, CMIP5 climate model output data require some
pre-processing. Data will need to be saved in a specific format, with
files stored in a specific directory structure. Full details of the
required file and directory structure are provided in the package's
``futureheatwaves'' vignette, while tips and an R script for conducting
this processing starting from CMIP5 netCDF files are given in the
``starting\_from\_netcdf'' vignette.

@@

\noindent However, this is not the only accepted heat wave definition in
the scientific literature. A variety of different heat wave definitions
have been used to identify heat waves in a time series of temperature
data
\citep{smith2013heat, kent2014heat, chen2015influence, anderson2009weather},
and the choice of heat wave definitions can influence both projected
heat wave trends \citep{smith2013heat} and estimates of health risks
during events
\citep{chen2015influence, kent2014heat, anderson2009weather}. Further,
other types of extreme events will be defined differently than heat
waves (for example, frost day spells may be defined as one or more days
with temperature at or below 32\textsuperscript{o}F /
0\textsuperscript{o}C).

Therefore, this package allows the user to extensively customize the
definition used to identify extreme events. Users can write a custom R
function with either a different heat wave definition (see
\citet{smith2013heat} and \citet{kent2014heat} for listings of some of
the definitions used in scientific studies) or with a definition
appropriate for a different type of extreme event (e.g., one or more
days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for
frost day spells). For heat wave identification, researchers might want
to use a specific definition, for example, because it matches the
definition used by local health officials to declare heat wave warnings
or, in the case of health impact assessments, to match with a definition
used in an epidemiological study. For studies of other extreme events, a
heat wave definition likely will not be applicable and so a customized
definition is necessary.

Three components of the extreme event definition can be easily
customized in the \code{gen\_hw\_set} function call, without creating a
new R function to use to identify heat waves. First, many extreme event
definitions are based on conditions that are rare in the study location
\citep{IPCCch1}, but may vary in how rare conditions must be for a
period to qualify as an extreme event. For example, some of the
different definitions used to identify heat waves vary only in the
percentile temperature used for a threshold (e.g., one definition is
\(\ge2\) days at or above the 98\textsuperscript{th} percentile
temperature at a location while another is \(\ge2\) days at or above the
99\textsuperscript{th} percentile temperature;
\citet{kent2014heat, smith2013heat}). Therefore, the
\pkg{futureheatwaves} package allows users to change the percentile of
the variable of interest required for an extreme event using the
\code{probThreshold} option in \code{gen\_hw\_set}. Other heat wave
definitions vary only in the number of consecutive days that must be
over the threshold for a period to quality as an extreme event (e.g.,
one definition is \(\ge2\) days at or above the 98\textsuperscript{th}
percentile temperature at a location while another is \(\ge4\) days at
or above the 98\textsuperscript{th} percentile temperature;
\citet{anderson2009weather}). Therefore, the package allows the user to
change the number of days used in the heat wave definition using the
\code{numDays} argument in the \code{gen\_hw\_set} function. Combined,
these two customization choices allow the user to identify heat waves
using many of the heat wave definitions used in previous climate and
health research-- for example, 9 of the 16 heat wave definitions used in
\citet{kent2014heat} could be fit using different combinations of these
two options for specifying threshold percentile and number of days.
Third, some extreme events like cold waves and frost day spells are
defined as a certain number of days below, rather than above, a
threshold. While the default is to identify events by searching for days
above a threshold, this behavior can be changed with the
\code{above\_threshold = FALSE} argument in the
\code{gen\_hw\_set function}.

Beyond these simpler options, the customization of the event definition
is even more extensive as one has the option of writing and using a
custom R function to identify extreme events. This functionality allows
the user to use definitions that either require a number of days above
an absolute threshold (e.g., maximum temperature of \(\ge\)
95\textsuperscript{o}F for \(\ge1\) day
\citet{kent2014heat, tan2007heat}; minimum temperature \(\le\)
0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or that
require a combination of thresholds to be met (e.g., maximum daily
temperature above a lower threshold every day of the heat wave and above
a higher threshold for a certain number of days within the heat wave;
\citet{kent2014heat, peng2011toward}). To use a customized event
definition, the user must write and load an R function that implements
the definition. This custom function is passed to the
\code{gen\_hw\_set} function using the \code{IDheatwavesFunction}
argument. To work correctly, this custom function must allow only
specific inputs and generate only specific outputs; details about the
required structure are provided in the \pkg{futureheatwaves} package
vignette. To increase processing speed when identifying extreme events,
we coded parts of the default event definition function in C++ and
synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users
should consider a similar strategy for custom heat wave definitions,
especially if processing a large number of climate projection files.

\subsection{Exploring adaptation
scenarios}\label{exploring-adaptation-scenarios}

Extreme events are often based on conditions that are rare for a
specific location using location-specific relative thresholds. These
thresholds are often defined for climate impact studies based on a
variable's distribution at that location in present-day or historical
data. However, it can be interesting to explore trends in extreme events
under climate change if extreme events are identified based on variable
distributions during the projection period, or another future period.
For some extreme events, impacts are associated with how rare the
conditions during the event are compared to norms in the location
\citep{anderson2009weather}. This relationship raises the question of
whether extreme events should be defined using a percentile threshold
based on present-day variable distributions or based on distributions in
the time period being projected. The \pkg{futureheatwaves} package
allows users to specify the time period to use when determining a
relative threshold for an event definition using the
\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set},
allowing users to explore how sensitive projections of impacts are to
this choice of the time period to use when determining relative variable
measures, including thresholds used for percentile-based event
definitions.

Similarly, some of the event characteristics (e.g.,
\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are also
calculated by the package based on relative temperature, providing
measures of how the value of the variable of interest during an extreme
event compares to the typical distribution of that variable at that
location (e.g., the ``mean.var.quantile'', ``min.var.quantile'', and
``max.var.quantile'' characteristics, Table \ref{tab:hwcharacteristics})
or how long conditions of a certain rarity persisted during the event
(e.g., the ``days.above.99th'' and ``days.above.99.5th''
characteristics, Table \ref{tab:hwcharacteristics}). These
characteristics are measured for each of the extreme events identified
by the \code{gen\_hw\_set} function by taking the absolute value of the
variable during the event (e.g., average temperature during the heat
wave is 90\textsuperscript{o}F ~32.2\textsuperscript{o}C) and comparing
it to the location's typical variable distribution. This process
generates relative measures of how intense the event is compared to what
is normal in that location (e.g., 90\textsuperscript{o}F
~32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of
year-round temperatures in the location).
